---
title: "Model description"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{model_description}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = FALSE}
library(PandemicsR)
```

# Overview

This package provides a framework for simulating the co-evolution of social
structure, opinion dynamics, and epidemic spreading in dynamic
multi-membership networks.

The model is designed to study how social organization, opinion polarization,
and epidemic spreading interact and influence each other.

The model combines three interacting processes:

1. **Schelling model: a bipartite group membership dynamics**
2. **Voters model: opinion dynamics model with 2 or 4 opinions**
3. **Epidemic model: SIR epidemic spreading process**

The three processes are coupled and evolve simultaneously in continuous time
using a Gillespie stochastic simulation algorithm.

The main assumption is that social organization, individual opinions, and
epidemic states are not independent. Instead, they continuously influence each
other during the simulation.

# Model assumptions

The model relies on several assumptions:

1. Individuals interact primarily through shared group memberships.
2. Individuals may belong to multiple groups simultaneously.
3. Opinion updates occur only inside groups.
4. Opinions are synchronized across all memberships of an individual.
5. Epidemic transmission follows a mean-field prevalence approximation.
6. Social and epidemic processes evolve on the same continuous-time scale.
7. All transitions are memoryless stochastic events.

# Population and network representation
The model therefore contains two coupled representations: the bipartite graph and random intersection graph. 


## Individuals and groups
The population consists of:

- $n$ individuals,
- $m$ social groups.

Individuals can belong to multiple groups simultaneously.

Memberships are represented as a bipartite network:

\[
B=(U,V,E)
\]

where:

- $U$ represents individuals,
- $V$ represents groups,
- $E$ represents membership relationships.

An edge between an individual and a group indicates that the individual
belongs to that group.


## Initial membership network

The initial bipartite graph is generated using a random intersection
construction.

The parameter $\lambda$ controls the expected density of memberships and
therefore influences:

- the average number of groups per individual,
- group sizes,
- social connectivity.

The model allows multi-membership behavior, meaning that individuals may
participate in several groups simultaneously.

## Projected social network

The bipartite network generates a projected Random Intersection Graph (RIG), which represents (potential) social interactions.

Two individuals are connected in the RIG (social layer) when they share at least one group in the bipartite graph (membership network). 

Because memberships evolve during the simulation, the projected social network
changes dynamically.

# Opinion dynamics

Each individual holds an opinion variable.

The model supports two configurations.

## Binary opinion model

The binary model contains two moderate states:

\[
o_i \in \{-1,+1\}
\]

where:

- $-1$ represents one opinion camp,
- $+1$ represents the opposing camp.

## Four-state opinion model

The extended model introduces radicalized states:

\[
o_i \in \{-2,-1,+1,+2\}
\]

where:

- $-1$ and $+1$ represent moderate opinions,
- $-2$ and $+2$ represent extreme opinions.

An individual's opinion is consistent across all groups to which they belong.

# Voter interaction dynamics

Opinion updates occur only through interactions inside groups.

For group $i$ define:

- $R_i$: number of moderate individuals with Red opinions,
- $B_i$: number of moderate individuals with Blue opinions.

If both camps exist in a group, voter conversion occurs at rate:

\[
r_{voter,i}
=
\gamma
\frac{R_iB_i}{R_i+B_i}
\]

where:

- $\gamma$ is the voter interaction parameter.

During an event, one individual is selected and changes opinion:

- a moderate Red individual may become Blue,
- a moderate Blue individual may become Red.

The interaction rate depends on the composition of the group rather than
absolute group size.

## Radicalization dynamics

When the four-state model is enabled, moderates may become extremists through local reinforcement inside groups. The interaction depends on the fractions of extremes and moderates within the group.

For group $i$:

- $E_{R,i}$ is the number of Red extremists,
- $E_{B,i}$ is the number of Blue extremists.

Radicalization rates are:

\[
r_{rad,R,i}
=
\alpha
\frac{E_{R,i}R_i}{m_i}
\]

and

\[
r_{rad,B,i}
=
\alpha
\frac{E_{B,i}B_i}{m_i}
\]

where:

- $\alpha$ controls radicalization strength,
- $m_i$ is the group size.

Extremists therefore influence moderate individuals from the same opinion
camp.

Optional deradicalization allows extremists to return to moderate states:

\[
-2 \rightarrow -1
\]

and

\[
+2 \rightarrow +1
\]

with rate controlled by the deradicalization parameter. Spontaneous radicalization is also implemented.

# Schelling membership dynamics

The social structure evolves through joining and leaving processes.

## Joining events

Individuals outside a group may join that group.

The joining rate depends on:

- the joining parameter,
- the availability of outsiders,
- population size and/or the group size.

A general formulation is:

\[
r_{join}
=
\frac{c}{m}
\frac{N_{outside}}{n}
\]

where:

- $c$ controls group formation,
- $N_{outside}$ is the number of individuals not belonging to the group.

Note: The rate can be scale by population size n and/or the group size m.

## Leaving events

Individuals may leave groups depending on local satisfaction.

For a group with tolerance threshold $T$, the leaving rate depends on the
fraction of similar individuals.

Individuals are more likely to leave groups where their opinion is
underrepresented.

For example, a Red opinion individual may leave according to:

\[
r_{leave}
=
\begin{cases}
\beta_{+}R_i, & \text{if similarity is below tolerance}\\
\beta_{-}R_i, & \text{otherwise}
\end{cases}
\]

The same mechanism applies symmetrically for blue individuals.

This implements a Schelling-like segregation mechanism where individuals are more likely to abandon groups in which their own opinion is underrepresented. Extreme individuals may also leave groups using analogous rules.

As memberships change, the underlying bipartite graph and the projected random intersection graph evolve over time.


# Epidemic model

The epidemic layer follows a stochastic SIR model.

Each individual has one epidemiological state:

\[
e_i \in \{S,I,R\}
\]

where:

- $S$ = susceptible,
- $I$ = infected,
- $R$ = recovered.

Initially, a set of individuals $I_0$ is randomly selected as infected.

## Infection dynamics

Transmission follows a mean-field prevalence model.

The infected prevalence is:

\[
p_I=\frac{I}{n}
\]

A susceptible individual becomes infected at rate:

\[
r_{infection}
=
\beta
\frac{I}{n}
\]

where:

- $\beta$ is the transmission parameter.

The model allows opinion-dependent infection rates:

\[
\beta_{negative}
\]

and

\[
\beta_{positive}
\]

Hence, epidemic susceptibility depends on opinion.

## Recovery dynamics

Recovery occurs independently at a fixed rate for infected individuals, regardless of opinion or membership status.

The recovery transition is:

\[
I \rightarrow R
\]

with rate:

\[
\gamma_{epi}
\]

where $\gamma_{epi}$ controls the recovery process.

# Gillespie simulation

All processes are integrated using a continuous-time Gillespie algorithm.

At each Gillespie step:
1. All event rates (social and epidemic) are computed:
- voter updates,
- joining events,
- leaving events,
- radicalization events,
- infection events,
- recovery events.

2. The total rate is \lambda_{\text{tot}} is computed by summing all enabled processes 
\[
\lambda_{\text{tot} =
\sum_k r_k
\]

3. The waiting time until the next event is sampled:

\[
\Delta t \sim Exp(\lambda_{\text{tot})
\]

4. One event is sampled according to its relative rate  \lambda_{\text{i} to the total rate \lambda_{\text{tot}. The selected event updates the system state.

Because event rates depend on the current state of the system, \lambda_{\text{i} to and \lambda_{\text{tot} change after every event.

# Stopping conditions

The simulation terminates when one of the following conditions is reached:

- maximum simulation time,
- epidemic extinction ($I=0$),
- complete polarization when the radicalization model is enabled.

